|

### 2023-01-19 text/ManuscriptDraft.Rmd
Reduced text manuscript version initiated.

---
new_session: FALSE
#title: |
#  **Towards individual-based pollination ecology: Automatic tracking of life histories of individual flowers**
#date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  word_document: default
  pdf_document:
    citation_package: natbib
    fig_caption: yes
    keep_tex: yes
    latex_engine: pdflatex
  html_document:
    df_print: paged
fontfamily: mathpazo
fontsize: 12pt
geometry: margin = 1in
header-includes:
- \usepackage{setspace}\doublespacing
- \usepackage[left]{lineno}
- \linenumbers
- \usepackage{dcolumn}
- \usepackage{caption}
- \usepackage{float}
- \usepackage{afterpage}
- \usepackage{siunitx}
- \usepackage{amsmath}
#keywords: Tracking, flowering phenology, Arctic
bibliography: ./library.bib
csl: ./journal-of-ecology.csl
---

```{r eval = FALSE, echo=FALSE}
library(knitr)
knit('ManuscriptDraft.Rmd')
rmarkdown::render("ManuscriptDraft.Rmd", output_file = "ManuscriptDraft_Resubmission.docx")
system2("open","ManuscriptDraft_Resubmission.docx")
```

```{r echo=FALSE}
# Function for creating word comments in Rmarkdown
# Use: This text contains a `r word_comment("This is the comment", "comment.")`.

@@

**Towards individual-based pollination ecology: Automatic tracking of life histories of individual flowers**

@@

Monitoring floral phenology at high temporal resolution is time-consuming, particularly in logistically challenging environments such as the Arctic. Consequently, flowering phenology of a population is often characterized simply by the date for onset of flowering, typically derived from weekly observations in permanent plots [@Prevey2021]. Such crude estimates may fail to reveal important dynamics in flowering phenology through the season. For example, different phenological factors, such as onset, peak, and end of flowering may respond differently to climate change [@cook2012] and estimation of the degree of mismatch between species may thus be sensitive to the choice of metric [@inouye2019]. The optimal metric for phenology depends on the research question, but often the choice of recording first flowering dates to represent flowering phenology is pragmatic rather than ideal. In fact, first flowering dates may be particularly ill suited for estimating sensitivity to climate change [@miller2008] since they represent the very extreme of the phenological distribution of a population and may be confounded with e.g. the size of the population considered. An accurate estimate of the phenological distribution of a population over the full season would allow for, for example, accurate quantification of resource availability through the season and temporal mismatches between species [@ramakers2020]. However, even detailed information on phenology of a population is still an aggregation of the phenology of the individual members of the population for which important variation may be masked [@inouye2019]. With the full distribution of flowering phenology at the level of individual flowers, it is possible to extract relevant characteristics of the distribution and test their individual sensitivities towards environmental factors, but acquiring such data requires efficient and preferably automated methods.

Cameras have become an ubiquitous tool for non-invasive and efficient monitoring of natural processes [@howe2017; @glover2019], but the implementation of image-based data collection has outpaced the development of methods for processing and analysing the big data output [@weinstein2018; @farley2018]. Gradually, computer vision and deep learning tools are being developed and applied for automatic processing and extraction of relevant data from images [@willi2019; @tabak2019; @hoye2021]. Indeed, automatic image-based monitoring of flowering phenology coupled with deep learning has facilitated collection of phenology data for specific species at very high temporal resolution [@mann2022]. While automatic detection and counting of flowers in image-series, as demonstrated in @mann2022, can produce detailed information about floral biology, phenological responses at the level of individual flowers may be indiscernible regardless of the temporal resolution of the data at population level. For example, a shortening of individual flower longevity may not be directly obvious at the population level. In fact, many research questions can only be directly explored on the basis of phenology data at the level of individual flowers and such information is critical for investigating the association between reproductive success and timing of flowering and/or flower longevity. Individual level data would also aid the understanding of whether flower visitation rates and/or seed set depends on the timing of flowering for the individual flower. Thus, automated recording of floral phenology by tracking the presence of individual flowers in image-series of permanent plots would be ideal and the feasibility of this as a second step subsequent of flower detection was explored in this study.

@@

Most multiple object tracking literature focus on tracking of humans [@luo2021; @milan2016], and these methods are not necessarily directly transferable to other use cases. Many tracking algorithms, such as the Simple online and real-time tracking (SORT) [@sort], rely on Kalmann filtering to predict the future location of an object by estimating its velocity from previous frames [@kalmann]. However, Kalman filtering assumes linear motion of the object and does not handle abrupt motion well. Flowers are constrained in movement by their stalk and violate this assumption. Other tracking approaches, such as DeepSORT [@deepsort2017], use tracking-by-detection in which convolutional neural networks (CNNs) are applied to continuously distinguish individuals through time, thus aiding the tracking algorithm. Individual flowers, however, are unlikely to be clearly distinguishable and one flower may look less alike between two frames than two flowers in a single frame, which prevents this approach. Finally, some tracking methods require manual initiation of tracks for a fixed number of objects in the first frame and the algorithm then tracks these objects in the subsequent frames [@hu2012]. These methods do not allow objects to enter or disappear from frame during tracking. In full-season monitoring data, flowers will disappear from the frame when they wilt, in which case their track should be terminated. Oppositely, new flowers should continuously be tracked as they bloom throughout the season. Additionally, a flower tracking method must handle flowers disappearing and appearing at the edges of the frame. Clearly, individual flower tracking is a challenging task.

Here, we present and evaluate a simple multiple object tracking algorithm for automatic tracking of individual flowers in image-series. Our tracking algorithm is based on centroid tracking and assigns individual objects to tracks based on the minimum distance between centroids of bounding boxes and does not rely on prediction of future positions. Further, the method allows both initiation and termination of tracks during tracking. We describe three user-defined parameters incorporated into the algorithm, demonstrate their effect on simulated data, and discuss why they may help optimise tracking of flowers in particular. We apply the algorithm on annotations of real flowers in time-lapse image-series from full-season monitoring of two widespread Arctic flowering plant species *Dryas integrifolia* and *D. octopetala*. For complex scenes with many flowers in close vicinity to each other, there is an increased risk of tracking errors regardless of the tracking method being used. For such cases, we suggest a conservative filtering approach based on the distance between flower tracks within the frame in order to identify tracks that are sufficiently spatially isolated and likely to be tracked with certainty. Overall, we show that information on floral biology can be automatically derived from image-based monitoring of flower plots.

@@

@@

To demonstrate our tracking algorithm and the effect of the three parameters, we used simulated data mimicking flower detections.To produce simulated detections of a flower, we defined a circle with radius rF and center point cF to delimit the boundaries of the flower. The first data point for the flower was then generated as a random point within this circle. Then, remaining data points (i.e. data points for the subsequent frames) were generated iteratively whith each new point being constrained to be within a circle with a center point of the position of the previously generated point and a radius rM, as well as limited to be within the boundaries of the flower. This approach allowed us to simulate specific scenarios on which to demonstrate our algorithm by controlling the number of flowers generated and the number of data points produced per flower, the boundaries of a flower, and the degree of movement between frames (i.e. the maximum distance between current and next point).

We generated four simulated data sets (Fig. 2) to demonstrate each of the three tracking parameters as well as the influence of frame rate on tracking reliability. In simulation A (Fig. 2A), we produced two flowers that are spatially distinct and temporally separated (i.e., one flower appears subsequent of the other). Each flower consists of ten data points and thus the data set spans over 20 frames. In simulation B, we produced four flowers that are spatially separated. Each flower consists of 5 data points, but the data set spans over 10 frames and each flower contains gaps of 1-3 data points at random temporal location. In simulation C, we produced four flowers that are in close proximity to each other and with a large degree of movement between frames. Each flower consist of 500 data points and the data sets spans over 500 frames. Finally, in simulation D, we produced four flowers consisting of a very high number of data points (10,000 per flower) but with only little movement between frames. Here, we used the same center point (cF) for all four flowers to increase overlap. Two pairs of flowers have a high degree of spatial overlap.

For each of the simulations, we present the tracking performance of our algorithm without the use of the three parameters and subsequently show how the parameters can be used to mitigate various challenges in the data and increase tracking accuracy. The two objects in simulation A are adjacent in time but spatially separated. We therefore introduce a limit to the spatial distance that is accepted between points in adjacent frames before they are considered unique objects. We do that with the max distance parameter. The objects in simulation B overlap in time but are well separated in space. However, each track contains gaps which can lead to track fragmentation. To allow gaps in track, we set a number of frames that a track can disappear before initiating a new track for points appearing in the same location. In simulation c, each object is allowed to move a lot between frames. Each object has a large spread of points and the four objects are in close proximity to each other. This means that between two frames, there is a substantial risk that a point for one object is closer to a point from another object rather than the point for that object, which will lead to identity shifts. To mitigate this, we calculate distances based on a current point and a set number of previous points in a track using the running mean parameter. This reduces the weighting on the displacement of points between the current and the previous frame. Finally, in simulation D, we have four objects that overlap in pairs of two, but the movement allowed between frames is very small, which means that a given point from an object is likely to be closest to the point from the same object in the previous frame. Effectively, while the overlap is substantial, the frame rate is sufficiently high that accurate tracking is ensured. To demonstrate the importance of frame rate for reliable tracking, we simulate a reduction in frame rate (i.e., increased degree of movement between frames) by basing tracking on every 10th, 20th, and 30th frame, respectively, and present these results.

![Figure 2: Simulated flower detection data for demonstration of our tracking algorithm and the effect of the three tracking parameters as well as the significance of frame rate for reliable tracking. A: 10 points per flower. rF = 200, rM = 100. Arrow shows the last point of flower a and the first point of flower b. B: 10 points per flower. rF = 200, rM = 100 C: 500 points per flower. rF = 200, rM = 200. D: 10,000 points per flower. rF = 100, rM = 1.](../figures/simulatedData.png){width="30%"}

When tracking moving objects, two fundamental errors can occur. First, a single trajectory may be split into several tracks, an error known as fragmentation. This can happen when an object is temporarily occluded or have disappeared from view and a new track is erroneously initiated when it reappears, or when identity shifts occur between tracks, for example when they switch positions between frames. Fragmentation is especially prone in settings with high densities of objects. Second, merging occurs when several trajectories are assigned the same track identity. This can occur when a track is not correctly terminated at the end of a trajectory but rather continues on the trajectory of another object. Here, importantly, the two individual trajectories can be tracked without internal errors (i.e. without mismatches), but the error lies in the fact that the two trajectories have not been assigned unique tracking identities but instead are tracked as a single object.

The optimal way of quantifying tracking performance depends on the goal of the tracking. For example, in our case of tracking flowers, associating information obtained in the images to the individual flower, for example flower visits, would require the track to be as correct as possible (i.e., identity shifts should be minimized). On the other hand, to derive flowering length in practice just requires correct association between the first and the last image of a flower while intermediate points can be ignored. Here, track fragmentation can lead to underestimation of flowering duration (and overestimation of number of flowers) while track merging can lead to overestimation of flowering duration (and underestimation of number of flowers). In some scenarios one may be interested in the number of individual flowers that existed in a plot, in which case the number of tracks obtained by automatic tracking should closely resemble the actual number of individuals in the series.

Here, we apply the commonly used multiple object tracking accuracy (MOTA) score which quantifies tracking performance based on tracking mismatches derived by counting the number of identity shifts within each track [@bernardin2008]. Mismatches occur when objects swap track identity because they are in close vicinity of each other or when an object periodically disappears and is assigned a new track identity when it reappears. Only the shifts in tracking identity are counted as mismatches in the MOTA score, while the number of points assigned to each track is not considered. For example, a ground truth track that was assigned one ID for the first half of its lifetime and another for the remaining half would only constitute one error.

where fn, fp, and mm are the number of false negatives, false positives, and track mismatches, respectively, for frame t divided by the total number of ground truth annotations across all frames. The MOTA score ranges from -Inf to 1 with 1 being perfect detection and tracking. Tracking accuracy is commonly evaluated between manual annotations and detections produced, for example, by a convolutional neural network (CNN). Therefore, false positives and false negatives are included as errors in the MOTA score. In our case, we are not tracking on data produced by a CNN but rather on manually annotated tracks and do not have false positives or false negatives. Thus, these variables are ignored in the calculation of the MOTA score.

To evaluate tracking accuracy of our algorithm, we calculate the total number of mismatches as well as the MOTA score (number of mismatches in relation to the total number of points). Additionally, we derive the number of tracks that were tracked with more than 80% accuracy. Finally, we compare the number of tracks identified by the automatic tracking with the true number of flowers in a series. These should ideally be equal. However, we note that a correct number of tracks (i.e., number of tracks equal to number of objects) does not necessarily translate to correct tracking, since individual tracks can be erroneously split (fragmentation) by the tracking algorithm while several tracks can be merged into one. We present the tracking performance without the use of the three tracking parameters along with the best performance (lowest number of mismatches/highest MOTA), and the best performance where the number of tracks correspond to the number of individual flowers in the series.

@@

This study is based on five full-season time-lapse image series collected at the following three Arctic sites (site abbreviation and sampling years in parentheses): Narsarsuaq (NARS), South Greenland (2018, 2019), Thule (THUL), North-West Greenland (2018), Ny-Ålesund (NYAA), Svalbard (2019). The images were captured using time-lapse cameras (Moultrie Wingscapes TimelapseCam Pro, Moultrie Products, Birmingham Alabama, USA) permanently positioned approx. 60 cm above areas with either *Dryas integrifolia* in Narsarsuaq and Thule or *D. octopetala* in Ny-Ålesund. The white flowers of these two species are insect-pollinated and held erect above cushions of foliage. The species are native to and widespread in Asian, North American, and European Arctic and alpine regions. They are geographically separated except for a possible hybrid zone in North-East Greenland [@philipp2003].

@@

We built a framework for tracking, filtering, and evaluating tracking of objects in time-lapse image series. The tracking algorithm (suppl. 1) is based on centroid tracking [@rahman2020; @bjerge2021]. That is, objects with minimum displacement between frames are associated. Centroids are derived from bounding boxes around objects stemming from manual annotation or automatic detection of objects. The algorithm has a set of user-adjusted parameters that can optimize tracking accuracy. The parameters are particularly relevant for tracking of objects that are constrained to a specific area such as flowers. However, the tracking algorithm could be used to track any objects. The speed of the tracking algorithm depends on the computational power available as well as the number of objects that are being tracked, but the time for computation will be negligible in most use-cases. Tracking of a series containing 85 objects ran at \~0.02 seconds per frame on a standard laptop. This, together with the fact that the algorithm handles frames sequentially, means that it can be applied both offline (on a set of detections/annotations that have already been produced) or online (real-time tracking frame per frame).

Several issues complicate the task of tracking individual flowers through time-lapse image series. First, as flowers move, they may temporarily occlude the view of other flowers and flowers close to the edge of the frame may move in and out of view between images. This can cause a track to be lost and fragmentation to occur (Fig. 4A). However, if a flower reappears in the same area as a flower is already being tracked after disappearing in only a few frames, it is a reasonable assumption that it is the same individual and not that the old flower wilted/disappeared and a new one developed. The parameter **max gap** sets the number of frames a flower can be absent before reappearing without the track ending and a new initiating. If a given flower has not been detected in a number of frames exceeding the max gap size, the track will be terminated. Concurrently, this can deal with potential false negatives, as a track that is missing a bounding box will not be lost.

Second, a standard centroid tracking algorithm will associate a point with a track based on minimum distance only, disregarding the absolute distance between centroids, which may lead to tracking errors (Fig. 4B). However, as flowers are constrained in their displacement, we can assume a maximum distance between points and force the initiation of tracks for points that exceed this threshold. We set this threshold with the parameter **max distance,** measured as the euclidean distance in number of pixels between objects in neighbouring frames.

Third, reliable tracking based on distance requires that movement of individuals are captured at sufficient frequency, such that each tracks are not switched between frames. Stand-alone image-based monitoring of the relatively slow process of flower development through full growing seasons (several months) is done with low frame rate, limited by power consumption and storage capacity both on and off the camera. Here, we used several images per day and while this temporal resolution captures the phenology of the flowers at very high detail, they fail to capture the full movement of flowers. For example, as the wind shifts, the flower heads change direction, which in the image series happens instantaneously (i.e., between two consecutive frames) (Fig. 4C). The fact that the frame rate is insufficient to capture smooth movement means that bounding boxes around a single individual in two consecutive frames cannot be associated based on overlap (e.g. intersection over union) as there often is no overlap. With a high number of flowers in close vicinity to each other, there's a substantial risk of identity switching between flowers as they move among each other (Fig. 1C;4C). Additionally, as the flowers are constrained by their stalk, there is a limit to the distance they can move, which prevents prediction of future location of an individual based on its previous trajectory. Here, we introduce the tracking parameter **running mean**, with which we can base tracking on the distance between a point in the current frame and the running mean of the positions of a number of previous points in a track, taking advantage of the fact that a flower moves around a central point. Increasing the value for this parameter decreases the weighting of movement between individual frames. Setting the parameter to 1 means that tracking is based on the distance to the centroid in the previous frame only.

@@

![Figure 4: Centroid tracking may produce erroneous associations when objects move or disappear between frames. A: A track may be lost when a flower temporarily disappears from view. If gaps in tracks are not allowed during tracking, a new track will erroneously be initiated when the flower reappears. B: A flower may be linked to an incorrect track despite large distance between them. Here, a flower disappears from view and another reappears between two frames. If no threshold for maximum distance between an object and a track is set, the reappearing flower will be linked to the track of the disappearing flower based on shortest distance despite the absolute distance being large. C: A flower may be linked to and incorrect track despite shortest distance to the running mean of the correct track. As flowers move around a central point, it is relevant to base tracking on a running mean of previous positions. Dots show position of flower in previous frame. Crosses show running mean of positions in x \> 1 number of previous frames. Dots and crosses coloured according to track identity.](../figures/errorFigure.png){width="100%"}

#### Input data for the tracking algorithm

The input data for the tracking algorithm are bounding box coordinates for each object in each frame given in a commonly used format [@everingham2010]: frame number, x-min, y-min, x-max, y-max. Here, x-min, y-min is the top left corner of the bounding box and x-max, y-max is the bottom right corner. The bounding box centroid is calculated on the basis of these coordinates within the algorithm. Calculating tracking accuracy requires a ground truth ID for each bounding box which can be given as the sixth column. As this data returned from both manual object annotations and object detection methods using automatic object detection methods based on deep learning, the method can be applied on both manually and automatically derived data.

#### Identifying optimal user parameters

To explore the effect of the three tracking parameters and to identify the optimal combination of parameters for our use-case of tracking flowers, we applied the tracking algorithm on each of the five image series with every combination of a range of values for each parameter (3179 combinations): 0-160 with a step size of 10 for max gap; 1-160 with a step size of 10 for running mean; 0-1000 with a step size of 100 for max distance. Note that max distance set to zero ignores the parameter altogether. We identified the setting(s) that returned the most accurate tracks for each series and compare the tracking results between optimal settings and all parameters ignored (max gap = 0, running mean = 1, max distance = 0).

@@

>>>>

---
new_session: FALSE
#title: |
#  **Towards individual-based floral biology: Automatic tracking of life histories of individual flowers**
#date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  word_document: default
  pdf_document:
    citation_package: natbib
    fig_caption: yes
    keep_tex: yes
    latex_engine: pdflatex
  html_document:
    df_print: paged
fontfamily: mathpazo
fontsize: 12pt
geometry: margin = 1in
header-includes:
- \usepackage{setspace}\doublespacing
- \usepackage[left]{lineno}
- \linenumbers
- \usepackage{dcolumn}
- \usepackage{caption}
- \usepackage{float}
- \usepackage{afterpage}
- \usepackage{siunitx}
- \usepackage{amsmath}
#keywords: Tracking, flowering phenology, Arctic
bibliography: ./library.bib
csl: ./journal-of-ecology.csl
---

```{r eval = FALSE, echo=FALSE}
library(knitr)

knit('ManuscriptDraft.Rmd')
rmarkdown::render("ManuscriptDraft.Rmd", output_file = "ManuscriptDraft_Resubmission.docx")
system2("open","ManuscriptDraft_Resubmission.docx")
```
```{r echo=FALSE}
# Function for creating word comments in Rmarkdown
# Use: This text contains a `r word_comment("This is the comment", "comment.")`.

@@

**Towards individual-based floral biology: Automatic tracking of life histories of individual flowers**

@@

Monitoring floral phenology at high temporal resolution is time-consuming, particularly in logistically challenging environments such as the Arctic. Consequently, flowering phenology of a population is often characterized simply by the date for onset of flowering, typically derived from weekly observations in permanent plots [@Prevey2021]. Such crude estimates may fail to reveal important seasonal dynamics in flowering phenology.. For example, different phenological factors, such as onset, peak, and end of flowering may respond differently to climate change [@cook2012] and estimation of the degree of mismatch between species may thus be sensitive to the choice of metric [@inouye2019]. The optimal metric for phenology depends on the research question, but often the choice of recording first flowering dates to represent flowering phenology is pragmatic rather than ideal. In fact, first flowering dates may be particularly ill suited for estimating sensitivity to climate change since they represent the very extreme of the phenological distribution of a population and may be confounded with e.g. the size of the population considered [@miller2008]. An accurate estimate of the phenological distribution of a population over the full season would allow for, for example, accurate quantification of resource availability through the season and temporal mismatches between species [@ramakers2020]. However, even detailed information on phenology of a population is still an aggregation of the phenology of the individual members of the population for which important variation may be masked [@inouye2019]. With individual level floral phenology it is possible to extract relevant characteristics of the distribution and test their individual sensitivities towards environmental factors, but acquiring such data requires efficient and automated methods.

Cameras have become an ubiquitous tool for non-invasive and efficient monitoring of natural processes [@howe2017; @glover2019], but the implementation of image-based data collection has outpaced the development of methods for processing and analysing the big data output [@weinstein2018; @farley2018]. Gradually, computer vision and deep learning tools are being developed and applied for automatic extraction of relevant data from images [@willi2019; @tabak2019; @hoye2021]. Indeed, automatic image-based monitoring of flowering phenology coupled with deep learning has facilitated collection of phenology data for specific species at very high temporal resolution [@mann2022]. While automated detection and counting of flowers in image-series, as demonstrated in @mann2022, can produce detailed information about floral biology, phenological responses at the level of individual flowers may be indiscernible regardless of the temporal resolution of the data at population level. For example, a shortening of individual flower longevity may not be directly obvious at the population level. In fact, many research questions can only be directly explored on the basis of individual level floral phenology data and such information is critical for investigating the association between reproductive success and timing of flowering and/or flower longevity. Individual level data would also aid the understanding of whether flower visitation rates and/or seed set depends on the timing of flowering for the individual flower. Thus, automated recording of floral phenology by tracking the presence of individual flowers in image-series of permanent plots would be ideal and the feasibility of this as a step subsequent of flower detection was explored in this study.

@@

Most multiple object tracking (MOT) methods focus on tracking of humans [@luo2021; @milan2016], and are not necessarily directly transferable to other use cases. Many tracking algorithms, such as the Simple online and real-time tracking (SORT) [@sort], rely on Kalmann filtering to predict the future location of an object based on its previous velocity [@kalmann]. However, Kalman filtering assumes linear object motion and does not handle abrupt movement well. Flowers are constrained in movement by their stalk and violate this assumption. Other tracking approaches, such as DeepSORT [@deepsort2017], use tracking-by-detection in which convolutional neural networks (CNNs) are applied to continuously distinguish individuals through time, thus aiding the tracking algorithm. Individual flowers, however, are unlikely to be clearly distinguishable and one flower may look less alike between two frames than two flowers in a single frame, which prevents this approach. Finally, some algorithms require manual initiation of a fixed number of tracks in the first frame and do not allow objects to enter or disappear during tracking [@hu2012]. In full-season monitoring data, flowers should continuously be tracked as they bolt and wilt throughout the season Additionally, a flower tracking method must handle flowers disappearing and appearing at the edges of the frame. Clearly, individual flower tracking is a challenging task.

Here, we present and evaluate a simple MOT algorithm for automatic tracking of individual flowers in image-series. Our tracking algorithm is based on centroid tracking and assigns objects to tracks based on the minimum distance between centroids of bounding boxes, and allows initiation and termination of tracks during tracking. We describe three user-defined parameters incorporated into the algorithm, demonstrate their effect on simulated data, and discuss their applicability in flower tracking. For complex scenes with many flowers in close vicinity to each other, there is an increased risk of tracking errors regardless of the tracking method being used. For such cases, we suggest a conservative filtering approach based on the distance between flower tracks within the frame in order to identify tracks that are sufficiently spatially isolated and likely to be tracked with certainty. We apply the algorithm on annotations of real flowers in time-lapse image-series from full-season monitoring of two widespread Arctic flowering plant species *Dryas integrifolia* and *D. octopetala*. Overall, we show that information on floral biology can be automatically derived from image-based flower plot monitoring.

@@

In MOT, two fundamental errors can occur. First, a single trajectory may be split into several tracks, an error known as fragmentation. This can happen when an object is temporarily occluded or have disappeared from view and a new track is erroneously initiated when it reappears, or when identity shifts occur between tracks, for example when they switch positions between frames. Fragmentation is especially prone in settings with high densities of objects. Second, merging occurs when several trajectories are assigned the same track identity. This can occur when a track is not correctly terminated at the end of a trajectory but rather continues on the trajectory of another object. Here, importantly, the two individual trajectories can be tracked without internal errors (i.e. without mismatches).

The optimal approach for quantification of tracking performance depends on the goal of the tracking. For example, in our case of tracking flowers, associating information obtained in the images to the individual flower, for example flower visits, would require the track to be as correct as possible (i.e., identity shifts should be minimized). On the other hand, to derive flowering length in practice just requires correct association between the first and the last image of a flower while intermediate points can be ignored. Here, track fragmentation can lead to underestimation of flowering duration (and overestimation of number of flowers) while track merging can lead to overestimation of flowering duration (and underestimation of number of flowers). In some scenarios, one may be interested in the number of individual flowers that existed in a plot, in which case the number of tracks obtained by automatic tracking should closely resemble the actual number of individuals in the series.

Here, we apply the commonly used MOT accuracy score (MOTA) which quantifies tracking performance based on tracking mismatches derived by counting the number of identity shifts within each track [@bernardin2008]. Only shifts in tracking identity are counted as mismatches in the MOTA score, while the number of points assigned to each track is not considered.

where fn, fp, and mm are the number of false negatives, false positives, and track mismatches, respectively, for frame t divided by the total number of ground truth annotations across all frames. The MOTA score ranges from -Inf to 1, with 1 being perfect detection and tracking. In our case, we can ignore the variables false positives and false negatives, as we are tracking on manually annotated tracks that are considered the ground truth.

To evaluate tracking accuracy of our algorithm, we calculate the total number of mismatches as well as the MOTA score (number of mismatches in relation to the total number of points). Additionally, we derive the number of tracks that were tracked with more than 80% accuracy. Finally, we compare the number of tracks identified by the automatic tracking with the true number of flowers in a series. These should ideally be equal. However, we note that a correct number of tracks (i.e., number of tracks equal to number of objects) does not necessarily translate to correct tracking, since single tracks can be erroneously fragmented simultaneously with merging of multiple tracks. We present the tracking performance without the use of the three tracking parameters along with the best performance (lowest number of mismatches/highest MOTA), and the best performance where the number of tracks correspond to the number of individual flowers in the series.

@@

To demonstrate our tracking algorithm and the effect of the three parameters, we used simulated data mimicking flower detections.To produce simulated detections of a flower, we defined a circle with radius rF and center point cF to delimit the boundaries of the flower. The first data point for the flower was then generated as a random point within this circle. Then, remaining data points (i.e. data points for the subsequent frames) were generated iteratively with each new point being constrained to be within a circle with a center point of the position of the previously generated point and a radius rM, as well as limited to be within the boundaries of the flower. This approach allowed us to simulate specific scenarios on which to demonstrate our algorithm by controlling the number of flowers generated, their positions and boundaries, the number of data points produced per flower, and the degree of movement between frames (i.e. the maximum distance between current and next point).

We generated four data sets (Fig. 2) to demonstrate each tracking parameter as well as the influence of frame rate on tracking reliability. In simulation A (Fig. 2A), we produced two flowers that are spatially distinct and temporally separated (i.e., one flower appears subsequent of the other). Each flower consists of ten data points and thus the data set spans over 20 frames. In simulation B, we produced four flowers that are spatially separated. Each flower consists of 5 data points, but the data set spans over 10 frames and each flower contains random gaps of 1-3 data points. In simulation C, we produced four flowers that are in close proximity to each other and with a large degree of movement between frames. Each flower consists of 500 data points and the data set spans over 500 frames. Finally, in simulation D, we produced four flowers with very high number of data points (10,000 per flower) but with only little movement between frames. Here, we used the same center point (cF) for all four flowers to increase overlap. Two pairs of flowers have a high degree of spatial overlap.

For each of the simulations, we present the tracking performance of our algorithm without the use of the three parameters and subsequently show how the parameters can be used to mitigate various challenges in the data and increase tracking accuracy. The two objects in simulation A are adjacent in time but spatially separated. We therefore introduce a limit to the spatial distance that is accepted between points in adjacent frames before they are considered unique objects. We do that with the max distance parameter. The objects in simulation B overlap in time but are well separated in space. However, each track contains gaps which can lead to track fragmentation. To allow gaps in track, we set a number of frames that a track can disappear before initiating a new track for points appearing in the same location. In simulation c, each object is allowed to move a lot between frames. Each object has a large spread of points and the four objects are in close proximity to each other. This means that between two frames, there is a substantial risk that a point for one object is closer to a point from another object rather than the point for that object, which will lead to identity shifts. To mitigate this, we calculate displacements based on the current point and a set number of previous points in a track using the running mean parameter. This reduces the weighting on the displacement between the current and the previous point. Finally, in simulation D, we have four objects that overlap in pairs of two, but with little movement between frame. Therefore, a point from a given object is likely to be closest to the point from the same object in the previous frame. Effectively, while the overlap is substantial, the frame rate is sufficiently high that accurate tracking is ensured. To demonstrate the importance of frame rate for reliable tracking, we simulate a reduction in frame rate (i.e., increased degree of movement between frames) by basing tracking on every 10th, 20th, and 30th frame, respectively.

![Figure 2: Simulated flower detection data to demonstrate the tracking algorithm and the effect of the three tracking parameters, max distance, max gap, and running mean (refer to Automatic flower tracking section for definitions), as well as the significance of frame rate for reliable tracking. A: 10 points per flower, rF (radius of circle delimiting flower points) = 200, rM (radius of circles delimiting distance between consecutive points) = 100. Arrow shows the last point of flower a and the first point of flower b. B: 10 points per flower, rF = 200, rM = 100. C: 500 points per flower, rF = 200, rM = 200. D: 10,000 points per flower, rF = 100, rM = 1.](../figures/simulatedData.png){width="30%"}

@@

This study is based on five full-season time-lapse image series collected at the following three Arctic sites (site abbreviation and sampling years in parentheses): Narsarsuaq (NARS), South Greenland (2018, 2019), Thule (THUL), North-West Greenland (2018), Ny-Ålesund (NYAA), Svalbard (2019). The images were captured using time-lapse cameras (Moultrie Wingscapes TimelapseCam Pro, Moultrie Products, Birmingham Alabama, USA) permanently positioned approx. 60 cm above areas with either *D. integrifolia* in Narsarsuaq and Thule or *D. octopetala* in Ny-Ålesund. The white flowers of these two species are insect-pollinated and held erect above cushions of foliage. The species are native to and widespread in Asian, North American, and European Arctic and alpine regions. They are geographically separated except for a possible hybrid zone in North-East Greenland [@philipp2003].

@@

We present a framework for tracking, filtering, and evaluating tracking of objects in time-lapse image series. The tracking algorithm (suppl. 1) is based on centroid tracking [@rahman2020; @bjerge2021]. That is, objects with minimum displacement between frames are associated. Centroids are derived from bounding boxes around objects stemming from manual annotation or automatic detection of objects. Thus, the input data for the tracking algorithm are bounding box coordinates (top left corner and bottom right corner) for each object [@everingham2010]. The algorithm has a set of user-adjusted parameters that can optimize tracking accuracy. The parameters are particularly relevant for tracking of objects that are constrained to a specific area such as flowers.

Several issues complicate the task of tracking individual flowers through time-lapse image series. First, as flowers move, they may temporarily occlude the view of other flowers and flowers close to the edge of the frame may move in and out of view between images. This can cause a track to be lost and fragmentation to occur (Fig. 4A). However, if a flower reappears shortly after having disappeared from view, it is a reasonable assumption that it is the same individual. The parameter **max gap** sets the number of frames a flower can be absent before reappearing without triggering track termination and initiation of a new track. Concurrently, the parameter can prevent track termination owing to false negatives.

Second, a standard centroid tracking algorithm will associate a point with a track based on minimum distance only, disregarding the absolute distance between centroids, which may lead to tracking errors (Fig. 4B). However, as flowers are constrained in their displacement, we can assume a maximum distance between points and force the initiation of tracks for points that exceed this threshold. We set this threshold with the parameter **max distance,** measured as the euclidean distance in number of pixels between objects in neighboring frames.

Third, reliable tracking based on distance requires that movement of individuals are captured at sufficient frequency to prevent identity shifts. Stand-alone image-based monitoring of the relatively slow process of flower development through full growing seasons (several months) is likely to be done with low frame rate, limited by power consumption and storage capacity both on and off the camera. Here, we used several images per day and while this temporal resolution captures the phenology of the flowers at very high detail, they fail to capture the full movement of flowers. For example, as the wind shifts, the flower heads change direction, which in the image series happens instantaneously (i.e., between two consecutive frames) (Fig. 4C). The fact that the frame rate is insufficient to capture smooth movement means that bounding boxes around a single individual in two consecutive frames cannot be associated based on overlap (e.g. intersection over union) as there often is no overlap. With a high number of flowers in close vicinity to each other, there is a substantial risk of identity shifts between flowers (Fig. 1C;4C). Additionally, as the flowers are constrained by their stalk, there is a limit to the distance they can move, which prevents prediction of future location of an individual based on its previous trajectory. Here, we use the tracking parameter **running mean**, with which we can base tracking on the distance between a point in the current frame and the running mean of the positions of a number of previous points in a track, taking advantage of the fact that a flower moves around a central point. Increasing the value for this parameter decreases the weighting of movement between individual frames. Setting the parameter to 1 means that tracking is based on the distance to the centroid in the previous frame only.

@@

![Figure 4: Centroid tracking may produce erroneous associations when objects move or disappear between frames. A: A track may be lost when a flower temporarily disappears from view. If gaps in tracks are not allowed during tracking, a new track will erroneously be initiated when the flower reappears. B: A flower may be linked to an incorrect track despite large distance between them. Here, a flower disappears from view and another reappears between two frames. If no threshold for maximum distance between an object and a track is set, the reappearing flower will be linked to the track of the disappearing flower based on shortest distance despite the absolute distance being large. C: A flower may be linked to an incorrect track despite shortest distance to the running mean of the correct track. As flowers move around a central point, it is relevant to base tracking on a running mean of previous positions. Dots show position of flower in previous frame. Crosses show running mean of positions in x \> 1 number of previous frames. Flowers, dots and crosses coloured according to identity.](../figures/errorFigure.png){width="100%"}

#### Testing the use of the tracking parameters

To explore the effect of the three tracking parameters and to identify the optimal combination of parameters for our use-case of tracking flowers, we applied the tracking algorithm on each of the five image series with every combination of a range of values for each parameter (3179 combinations): 0-160 with a step size of 10 for max gap; 1-160 with a step size of 10 for running mean; 0-1000 with a step size of 100 for max distance. Note that max distance set to zero ignores the parameter altogether. We present the tracking accuracies for three different scenarios: Parameters zeroed (PZ) (Max gap = 0; Max distance = 0; Running mean = 1) (hereafter referred to as naive tracking), minimum number of mismatches achieved, and minimum number of mismatches achieved with a number of tracks equal to the number of flowers in the series.

@@

Owing to the limited resolution of the cameras used in the present study, it is unlikley that detailed information about anther opening or pollen release can be recorded from these images, but with increased image quality, this may be possible.
